Investigating Overfitting in Convolutional Neural Networks

### 2025-05-27 paper_formatted.tex
first draft for final paper; wrote reflection and changed formatting; fixed explanations and added on to some other sections.

% Packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[labelfont=it, textfont=it]{caption}
\usepackage{float}
\usepackage{tikz}
\usepackage{setspace}
\usepackage{times}
\usepackage[left=1in, right=1in, top=1in, bottom=1in]{geometry}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}

@@

    We trained 150 convolutional models, each on the same randomly selected small subset of 100 images
    from the Canadian Institute For Advanced Research - 10 dataset (CIFAR-10), which consists of 32x32 color images each labeled with one of ten classes (e.g., plane, boat, etc).
    % Each model had the same architecture, with the only difference between them being the number of parameters (\textit{see Figure 1 below}).
    The training dataset was constructed through stratified random sampling by randomly selecting 10 of each class of image, ensuring that it is representative of the entire dataset.
    The number of filters and neurons were varied between models in a way such that the sizes of the models (in parameter count) were roughly uniformly distributed, ranging from approximately 1m to 25m parameters.
    The architecture and the ratio of layer sizes were kept consistent between each model, ensuring that the only difference between them was the number of parameters (\textit{see Figure 1 below}).
    On the initialization of each model, the values of the parameters for that model were randomly set, so each model can be seen as randomly
    selected from all possible models of their respective size and architecture before training, ensuring the experiment is statistically valid.
    We controlled the parameter count by multiplying the amount of filters for the Conv2D layers and the amount of neurons
    for the hidden Dense layers by a scalar factor, $n$.

@@

    \section*{4. Data Display}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Scatter}
        \caption{Scatterplot of Parameter Count vs. Difference in Train and Test Accuracy (Train - Test)}
        \label{fig:scatterplot}
    \end{figure}
    There appears to be a moderate, positive, linear relationship between parameter count and difference between
    train and test accuracy (train - test). There appear to be a few possible high outliers above $x =$ 1.5 million parameters
    and a few possible low outliers above $x = 0.5$ million parameters.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid}
        \caption{Residual Plot (Parameter Count vs. Residuals)}
        \label{fig:residuals}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid Hist}
        \caption{Histogram of Residuals}
        \label{fig:residual_hist}
    \end{figure}
    \noindent The histogram of the residuals appears to be approximately symmetric, with a median around 0 and a range of
    $0.05-(-0.04)=0.09$. There do not appear to be any obvious outliers.

@@

    \subsubsection*{5.2.2 $t$-statistic}
    \noindent And now we can find the t-statistic, which represents how many standard deviations away our sample slope $b$ is from the population slope $\beta$, assuming that $\beta=0$:
    \begin{gather*}
        \mathrm{t} = \frac{b}{SE_\beta} \\[1em]
        \mathrm{t} = \frac{0.008312}{0.002008} = 4.139
    \end{gather*}

@@

    \noindent Based on the area under the curve in \textit{\autoref{fig:t_dist}} above $t=4.139$, our $p$-value is $2.916\cdot10^{-5}$.

@@

    The $p$-value of $2.916\cdot10^{-5}$ indicates that, assuming that the population slope $\beta=0$, there is around a
    $0.002916\%$ probability of getting a sample slope as great or greater than our sample slope of $0.008312$, in a sample of
    150 randomly selected models from the population of convolutional models of our structure and of size 1 million parameters to 25 million parameters, trained
    on a randomly selected subset of 100 images from the CIFAR-10 train dataset and tested on the full test dataset of CIFAR-10.
    Because the $p$-value of $2.916\cdot10^{-5}$ is less than $\alpha=0.05$, we reject the null hypothesis, which states that the population slope $\beta$
    of the least-squares regression line that relates parameter count to difference in train and test accuracy (train - test) is $0$,
    in the population defined above.
    The data do provide convincing evidence that the population slope $\beta$ is greater than $0$, and that there is a positive correlation between
    parameter count and difference between train and test accuracy (train - test) in the population defined above.

@@

>>>>

% Packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[labelfont=it, textfont=it]{caption}
\usepackage{float}
\usepackage{tikz}
\usepackage{setspace}
\usepackage{times}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}

@@

    We trained 150 convolutional models, each on the same randomly selected small subset of 100 images
    from the Canadian Institute For Advanced Research - 10 dataset (CIFAR-10), which consists of 32x32 color images each labeled with one of ten classes (e.g., plane, boat, etc).
    % Each model had the same architecture, with the only difference between them being the number of parameters (\textit{see Figure 1 below}).
    The training dataset was constructed through stratified random sampling by randomly selecting 10 of each class of image, ensuring that it is representative of the entire dataset.
    The number of filters and neurons were varied between models in a way such that the sizes of the models (in parameter count) were roughly uniformly distributed, ranging from approximately 1m to 25m parameters.
    The architecture and the ratio of layer sizes were kept consistent between each model, ensuring that the only difference between them was the number of parameters (\textit{see Figure 1 below}).
    On the initialization of each model, the values of the parameters for that model were randomly set, so each model can be seen as randomly
    selected from all possible models of their respective size and architecture before training, ensuring the experiment is statistically valid.
    All of the above precautions aimed to eliminate confounding with other variables, ensuring that any change in the accuracy of the models was actually caused by the change in parameter count.
    We controlled the parameter count by multiplying the amount of filters for the Conv2D layers and the amount of neurons
    for the hidden Dense layers by a scalar factor, $n$.

@@

    \section*{4. Data Display}
    \begin{figure}[H]
        \centering
        \includegraphics[width=1\textwidth]{Images/Scatter}
        \caption{Scatterplot of Parameter Count vs. Difference in Train and Test Accuracy (Train - Test)}
        \label{fig:scatterplot}
    \end{figure}

    \begin{table}[ht]
        \centering
        \begin{tabular}{|l|r|r|}
            \hline
            Predictor & Coef & SE Coef \\
            \hline
            Constant & 0.7454 &  \\
            Param Count (x10 mil)   & 0.0083 & 0.0020 \\
            \hline
            \multicolumn{2}{|l|}{$S = 0.01703$} & R-Sq = 10.37\% \\
            \hline
        \end{tabular}
        \caption{Regression Summary}
        \label{tab:minitab}
    \end{table}

    \noindent
    Based on the scatter plot (\textit{\autoref{fig:scatterplot}}) and the correlation coefficient of $r=\sqrt{0.1037}=0.3220$ (\textit{from \autoref{tab:minitab}}), there appears to be a weak,
    positive, linear relationship between parameter count and difference between train and test accuracy (train - test).
    There appear to be a few possible high outliers above $x =$ 1.5 million parameters
    and a few possible low outliers above $x = 0.5$ million parameters.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid}
        \caption{Residual Plot (Parameter Count vs. Residuals)}
        \label{fig:residuals}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid Hist}
        \caption{Histogram of Residuals}
        \label{fig:residual_hist}
    \end{figure}
    \noindent The histogram of the residuals appears to be approximately symmetric, with a median around 0 and a range of
    $0.05-(-0.04)=0.09$. There does not appear to be any strong skew or outliers.

@@

    \subsubsection*{5.2.2 $t$-statistic}
    \noindent And now we can find the t-statistic, which represents how many standard deviations away our sample slope $b$ is from the population slope $\beta$ on the sampling distribution, assuming that $\beta=0$:
    \begin{gather*}
        \mathrm{t} = \frac{b}{SE_\beta} \\[1em]
        \mathrm{t} = \frac{0.008312}{0.002008} = 4.139
    \end{gather*}

@@

    \noindent Based on the area under the curve in \textit{\autoref{fig:t_dist}} above, $t=4.139$, our $p$-value is $2.916\cdot10^{-5}$.

@@

    The $p$-value of $2.916\cdot10^{-5}$ indicates that, assuming that the population slope $\beta=0$, there is around a
    $0.002916\%$ probability of getting a sample slope as great or greater than our sample slope of $0.008312$, in a sample of
    150 randomly selected models from the population of convolutional models of our structure and of size 1 million parameters to 25 million parameters, trained
    on a randomly selected subset of 100 images from the CIFAR-10 train dataset and tested on the full test dataset of CIFAR-10.
    Because the $p$-value of $2.916\cdot10^{-5}$ is less than $\alpha=0.05$, we reject the null hypothesis, which states that the population slope $\beta$
    of the least-squares regression line that relates parameter count to difference in train and test accuracy (train - test) is $0$,
    in the population defined above.
    The data do provide convincing evidence that the population slope $\beta$ is greater than $0$, and that there is a positive correlation between
    parameter count and difference between train and test accuracy (train - test) in the population defined above. Thus, we can conclude that
    increasing the number of parameters in the model does indeed cause overfitting.

@@

    In our project, we aimed to determine whether increasing the parameter count of convolutional models of a specific structure caused the model to overfit,
    or essentially memorize the train dataset and perform well on images from that set but poorly on newly introduced images. We ensured equal representation
    of images of different classifications by randomly sampling by strata, with each strata being a group of images with the same classification.
    By keeping the structure of the models and the dataset on which they were trained on constant, as well as randomizing the parameters of the models
    on initialization, we prevented confounding with other variables, allowing us to establish a cause-and-effect relationship between parameter count
    and difference between train and test accuracy (train - test). Through a $t$-test for the population slope, we found that there is a positive relationship
    between parameter count and difference between train and test accuracy (train - test), and thus that increasing parameter count increases the difference and causes overfitting.

    We could have potentially made a Type I error, because we rejected the null hypothesis. If we had made a Type I error,
    that would mean that we found convincing evidence that increasing the parameter count of the model between 1m and 25m parameters
    caused overfitting, when in reality it didn't. A potential consequence of this would be that we would only use models with less than 1 million parameters,
    believing that models larger than 1 million parameters would perform worse on the test dataset, losing the potential
    of getting better accuracy by using larger models.

    However, our design included many limitations that could be addressed in future experiments.
    We only trained models with 1m to 25m parameters, which means we can only generalize to models of these sizes.
    In the future, we could conduct tests with a wider range of parameter sizes, which would allow us to generalize our results to a wider range of model sizes or potentially reveal a new trend with larger parameter sizes.

    Another major limitation was the compute time required. Despite training relatively simple models on a small dataset, it took over two hours to train all 150 models.
    If we wanted to scale the experiment up with more complex models or larger datasets, the compute time required could make the experiment impractical.
    In the future, we could optimize the experiment by training the models for less epochs or using less models, although this would decrease the power of the test.

    Finally, our results are only applicable to models of the specific architecture we used on a 100 image subset of the CIFAR-10 dataset, which is not useful to many people training neural networks.
    If we wanted the test to be more useful in helping people decide the size of their models, we could perform an experiment on different model architectures and datasets, allowing a broader generalization.

%    In the future, we could conduct tests beyond our current models architecture and our current parameter range,
%    to generalize to a larger population or find different trends. It is possible that as model size increases past
%    25 million parameters, that the difference in test and train accuracy could stop increasing and instead start decreasing.
%    We could also conduct tests with larger training datasets and see if the slope lowers as the training dataset size increases (because overfitting generally happens on small training datasets).
%    In the future we could also generalize our test to
    \clearpage

    \appendix
    \section*{Appendix: Raw Data}
    \addcontentsline{toc}{section}{Appendix: Raw Data}

    \begin{longtable}{@{}
        S[table-format=8.0]  c
        S[table-format=8.0]  c
        S[table-format=8.0]  c
        @{}}
        \caption{Raw Data: Parameter Count vs.\ Train–Test Accuracy Difference}\\
        \toprule
        \textbf{Param. Count} & \textbf{Accuracy}
        & \textbf{Param. Count} & \textbf{Accuracy}
        & \textbf{Param. Count} & \textbf{Accuracy} \\
        \midrule
        \endfirsthead

        \toprule
        \textbf{Param. Count} & \textbf{Accuracy}
        & \textbf{Param. Count} & \textbf{Accuracy}
        & \textbf{Param. Count} & \textbf{Accuracy} \\
        \midrule
        \endhead

        \bottomrule
        \endfoot

        \num{991995}   & 0.7642 & \num{1156636} & 0.7502 & \num{1314120} & 0.7201 \\
        \num{1481652}  & 0.7358 & \num{1653866} & 0.7463 & \num{1817930} & 0.7685 \\
        \num{1964108}  & 0.7325 & \num{2122007} & 0.7377 & \num{2279711} & 0.7495 \\
        \num{2443067}  & 0.7769 & \num{2612075} & 0.7425 & \num{2779778} & 0.7594 \\
        \num{2928578}  & 0.7481 & \num{3083068} & 0.7738 & \num{3247185} & 0.7501 \\
        \num{3409755}  & 0.7267 & \num{3574350} & 0.7405 & \num{3736748} & 0.7334 \\
        \num{3908963}  & 0.7632 & \num{4058757} & 0.7677 & \num{4231690} & 0.7548 \\
        \num{4387487}  & 0.7339 & \num{4537211} & 0.7427 & \num{4689447} & 0.7547 \\
        \num{4872926}  & 0.7316 & \num{5030648} & 0.7497 & \num{5190882} & 0.7655 \\
        \num{5353628}  & 0.7383 & \num{5518886} & 0.7448 & \num{5686656} & 0.7583 \\
        \num{5825435}  & 0.7481 & \num{5997767} & 0.7711 & \num{6162250} & 0.7160 \\
        \num{6304105}  & 0.7488 & \num{6472707} & 0.7484 & \num{6620711} & 0.7465 \\
        \num{6793468}  & 0.7498 & \num{6942372} & 0.7816 & \num{7119248} & 0.7650 \\
        \num{7274427}  & 0.7643 & \num{7455458} & 0.7449 & \num{7611411} & 0.7603 \\
        \num{7760207}  & 0.7178 & \num{7947148} & 0.7343 & \num{8108136} & 0.7494 \\
        \num{8261687}  & 0.7490 & \num{8413707} & 0.7760 & \num{8570111} & 0.7762 \\
        \num{8724930}  & 0.7676 & \num{8884187} & 0.7677 & \num{9041805} & 0.7389 \\
        \num{9203915}  & 0.7700 & \num{9364332} & 0.7457 & \num{9529295} & 0.7531 \\
        \num{9692511}  & 0.7650 & \num{9860327} & 0.7574 & \num{10026342} & 0.7423 \\
        \num{10183690} & 0.7185 & \num{10352393} & 0.7567 & \num{10480242} & 0.7178 \\
        \num{10654715} & 0.7686 & \num{10813533} & 0.7413 & \num{10990748} & 0.7619 \\
        \num{11122467} & 0.7582 & \num{11298745} & 0.7555 & \num{11465740} & 0.7503 \\
        \num{11644706} & 0.7431 & \num{11780277} & 0.7401 & \num{11950780} & 0.7537 \\
        \num{12133478} & 0.7462 & \num{12271857} & 0.7536 & \num{12445868} & 0.7624 \\
        \num{12586011} & 0.7623 & \num{12773485} & 0.7606 & \num{12951004} & 0.7709 \\
        \num{13093955} & 0.7385 & \num{13269953} & 0.7223 & \num{13399368} & 0.7681 \\
        \num{13544767} & 0.7861 & \num{13727548} & 0.7464 & \num{13874711} & 0.7629 \\
        \num{14055863} & 0.7519 & \num{14204772} & 0.7547 & \num{14391938} & 0.7496 \\
        \num{14526700} & 0.7440 & \num{14678075} & 0.7440 & \num{14864381} & 0.7778 \\
        \num{15017502} & 0.7522 & \num{15155155} & 0.8039 & \num{15348460} & 0.7520 \\
        \num{15504047} & 0.7453 & \num{15643907} & 0.7702 & \num{15836226} & 0.7647 \\
        \num{15977573} & 0.7830 & \num{16119548} & 0.7416 & \num{16278987} & 0.7585 \\
        \num{16422292} & 0.7620 & \num{16623490} & 0.7814 & \num{16768300} & 0.7512 \\
        \num{16930907} & 0.7684 & \num{17077047} & 0.7480 & \num{17277956} & 0.7671 \\
        \num{17425583} & 0.7586 & \num{17591340} & 0.7540 & \num{17740297} & 0.7658 \\
        \num{17889882} & 0.7834 & \num{18099850} & 0.7866 & \num{18250940} & 0.7443 \\
        \num{18402658} & 0.7526 & \num{18572987} & 0.7770 & \num{18726035} & 0.7677 \\
        \num{18879711} & 0.7298 & \num{19034015} & 0.7322 & \num{19246091} & 0.7612 \\
        \num{19344507} & 0.7307 & \num{19500695} & 0.7613 & \num{19657511} & 0.7302 \\
        \num{19814955} & 0.7481 & \num{19973027} & 0.7509 & \num{20131727} & 0.7725 \\
        \num{20291055} & 0.7952 & \num{20451011} & 0.7881 & \num{20670818} & 0.7372 \\
        \num{20832261} & 0.7392 & \num{20994332} & 0.7605 & \num{21157031} & 0.7665 \\
        \num{21320358} & 0.7831 & \num{21484313} & 0.7655 & \num{21648896} & 0.7812 \\
        \num{21814107} & 0.7388 & \num{21918875} & 0.7749 & \num{22085111} & 0.7238 \\
        \num{22251975} & 0.7380 & \num{22419467} & 0.7686 & \num{22567754} & 0.8055 \\
        \num{22736428} & 0.7818 & \num{22905730} & 0.7743 & \num{23008186} & 0.7285 \\
        \num{23178495} & 0.7603 & \num{23349432} & 0.7842 & \num{23520997} & 0.7418 \\
        \num{23693190} & 0.7992 & \num{23866011} & 0.7622 & \num{24018998} & 0.7663 \\
        \num{24193001} & 0.7622 & \num{24367632} & 0.7711 & \num{24542891} & 0.7671 \\
        \num{24654011} & 0.7606 & \num{24830295} & 0.7835 & \num{25007207} & 0.7820 \\
    \end{longtable}
